---
title: "Ratio Distributions"
author: "Susan Vanderplas"
---

This analysis considers the possible issues with `instructional_sch_4Y_share_growth` and `research_awards_growth_inc_nuf_fy20_fy24` variables. 
Both variables are formulated roughly as

$$\frac{X_{2024} - X_{2020}}{Y_{2024}-Y_{2020}},$$

where $X_{t}\sim f(x, \alpha)$ and $Y_t\sim f(y,\beta)$ -- that is, where we might expect that both X and Y have a common distribution function (whatever it might be), with independent and identically distributed observations taken on a yearly basis. 

That is, the goal is clearly to describe the percent of the university's growth attributable to a single department. 
However, these variables are ill formed, in ways that are demonstrable conceptually as well as mathematically. 

For one thing, there is no guarantee that either student credit hours or research awards have uniformly increased betwen 2020 and 2024. 
2020 is a tricky year for a lot of reasons, which makes it a problematic baseline. 
But, there are also factors within a department, like retirements in small departments, that might affect both research awards and credit hours generated -- these factors do not appear to be normalized by department size, which further complicates the problem.

Suppose there is a relatively simple university where there are three departments with 5, 15, and 35 individuals. 
Each individual teaches 4 3-credit courses per year, and course enrollment is randomly distributed between 5 (the minimum required to make) and 30 (the enrollment cap) students per course. 
We can describe this with a truncated normal distribution where the un-truncated distribution has mean=25, sd = 10, and we limit the possible outcomes to between 5 and 30, with any values rounded to the nearest whole number. 
In actuality, we should truncate the distribution at 4.5 and 30.5 to ensure that everything that would round up to 5 or down to 30 is included.  

```{r}
library(tidyverse)
library(truncdist)

class_size_dist <- tibble(x = seq(4.5, 30.5, .001), y = dtrunc(x, "norm", 4.5, 30.5, mean = 25, sd = 10)) |>
  mutate(rounded = round(x)) |>
  group_by(rounded) |>
  summarize(yint = sum(y*.001))

class_size_mean <- with(class_size_dist, sum(yint*rounded))
class_size_dist |>
  # select(rounded, yint) |>
  ggplot(aes(x = rounded, y = yint)) + geom_col() + 
  annotate(geom = "text", x = -Inf, y = Inf, label = sprintf("Mean = %.02f", class_size_mean), hjust = -0.1, vjust = 1.2) + 
  ylab("Density") + xlab("Class Size")

```


We will assume that our three departments are all equally popular and teach only the courses which they can staff with the department's members teaching a 2-2 load. 


```{r}
class_dist <- function(x) round(rtrunc(x, "norm", 4.5, 30.5, mean = 25, sd = 10))

if(!exists("data-analysis/simulation-ratio.Rdata")) {
  set.seed(20492803)

sch_results <- tibble(department = c("Small", "Med", "Large"), faculty = c(5, 15, 35)) |>
  crossing(year = c(2020:2024), iteration = 1:1000) |>
  mutate(faculty_id = map(faculty, ~tibble(faculty_id = 1:.))) |>
  unnest(faculty_id) |>
  mutate(course_id = map(faculty_id, ~tibble(faculty_course_id = 1:4))) |>
  unnest(course_id) |>
  mutate(enrollment = class_dist(n()), sch = 3*enrollment) 

save(sch_results, file = "data-analysis/simulation-ratio.Rdata")
} else {
  load("data-analysis/simulation-ratio.Rdata")
}

```


The simulation assumes that each year is approximately the same distribution. 
That is probably not completely accurate (enrollment changes, etc.) but is probably not totally off-base for this extremely simplified scenario where we also only have 3 departments. 

```{r}
sch_sum <- sch_results |> group_by(iteration, department, year) |> summarize(sch = sum(sch), enrollment = mean(enrollment)) 

ggplot(sch_sum, aes(x = year, y = sch, color = department, group = interaction(department, iteration))) + geom_line()
```



```{r}
overall_5y_sum <- sch_results |>
  group_by(iteration, year) |>
  summarize(sch = sum(sch), enrollment = mean(enrollment)) |>
  summarize(total_sch_2024_2020 = sch[5]-sch[1])


sch_5y_sum <- sch_sum |>
  left_join(overall_5y_sum) |>
  group_by(iteration, department) |>
  arrange(year) |>
  summarize(total_sch_2024_2020 = unique(total_sch_2024_2020), 
            sch_2024_2020 = sch[5]-sch[1], 
            share_growth_2024_2020 = sch_2024_2020/total_sch_2024_2020)


ggplot(sch_5y_sum, aes(x = department, y=share_growth_2024_2020, group = iteration)) + geom_line()


```


What do the ratios of department SCH to institutional SCH look like? 

