---
title: "Ratio Distributions"
author: "Susan Vanderplas"
---

This analysis considers the possible issues with `instructional_sch_4Y_share_growth` and `research_awards_growth_inc_nuf_fy20_fy24` variables. 
Both variables are formulated roughly as

$$\frac{X_{2024} - X_{2020}}{Y_{2024}-Y_{2020}},$$

where $X_{t}\sim f(x, \alpha)$ and $Y_t\sim f(y,\beta)$ -- that is, where we might expect that both X and Y have a common distribution function (whatever it might be), with independent and identically distributed observations taken on a yearly basis. 

That is, the goal is clearly to describe the percent of the university's growth attributable to a single department. 
However, these variables are ill formed, in ways that are demonstrable conceptually as well as mathematically. 

For one thing, there is no guarantee that either student credit hours or research awards have uniformly increased betwen 2020 and 2024. 
2020 is a tricky year for a lot of reasons, which makes it a problematic baseline. 
But, there are also factors within a department, like retirements in small departments, that might affect both research awards and credit hours generated -- these factors do not appear to be normalized by department size, which further complicates the problem.

Suppose there is a very small university where there are three departments with 5, 15, and 35 faculty members.  
Each individual teaches 4 3-credit courses per academic year, and course enrollment is randomly distributed between 5 (the minimum required to make) and 30 (the enrollment cap) students per course. 
We can describe this with a truncated normal distribution where the un-truncated distribution has mean=25, sd = 10, and we limit the possible outcomes to between 5 and 30, with any values rounded to the nearest whole number. 
In actuality, we should truncate the distribution at 4.5 and 30.5 to ensure that everything that would round up to 5 or down to 30 is included. 

```{r}
#| fig-cap: "Class enrollment distribution. Enrollment is generated according to a truncated normal distribution, where the untruncated distribution has mean 25 and SD 10; after truncation, the distribution has mean 20.72 and standard deviation 6.31."
#| label: fig-class-size-dist
#| fig-width: 6
#| fig-height: 4
library(tidyverse)
library(truncdist)

class_size_dist <- tibble(x = seq(4.5, 30.5, .001), y = dtrunc(x, "norm", 4.5, 30.5, mean = 25, sd = 10)) |>
  mutate(rounded = round(x)) |>
  group_by(rounded) |>
  summarize(yint = sum(y*.001))

class_size_mean <- with(class_size_dist, sum(yint*rounded))
class_size_dist |>
  # select(rounded, yint) |>
  ggplot(aes(x = rounded, y = yint)) + geom_col() + 
  annotate(geom = "text", x = -Inf, y = Inf, label = sprintf("Mean = %.02f", class_size_mean), hjust = -0.1, vjust = 1.2) + 
  ylab("Density") + xlab("Class Size")

```


We will assume that our three departments are all equally popular and teach only the courses which they can staff with the department's members teaching a 2-2 load. 
That is, this analysis excludes the possibility of teaching overloads and hiring adjuncts to teach additional courses. 
We also will ignore graduate students teaching courses. 


```{r}
class_dist <- function(x) round(rtrunc(x, "norm", 4.5, 30.5, mean = 25, sd = 10))

if(!file.exists("data-analysis/simulation-ratio.Rdata")) {
  set.seed(20492803)
  
  sch_results <- tibble(department = c("Small", "Med", "Large"), faculty = c(5, 15, 35)) |>
    crossing(year = c(2020:2024), iteration = 1:1000) |>
    mutate(faculty_id = map(faculty, ~tibble(faculty_id = 1:.))) |>
    unnest(faculty_id) |>
    mutate(course_id = map(faculty_id, ~tibble(faculty_course_id = 1:4))) |>
    unnest(course_id) |>
    mutate(enrollment = class_dist(n()), 
           sch = 3*enrollment) 
  
  save(sch_results, file = "data-analysis/simulation-ratio.Rdata")
} else {
  load("data-analysis/simulation-ratio.Rdata")
}

```


The simulation assumes that each year has approximately the same distribution (for now). 
That is probably not completely accurate (enrollment changes, etc.) but is probably not totally off-base for this extremely simplified scenario where we also only have 3 departments. 

```{r}
#| label: fig-sch-generation-time-dept
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Simulated SCH generated by each department over 5 years, given that class enrollment has distribution as in @fig-class-size-dist and the large, medium, and small departments have 35, 15, and 5 faculty respectively. Each faculty member teaches 4 courses per academic year."
sch_sum <- sch_results |> 
  group_by(iteration, department, year) |> 
  summarize(sch = sum(sch), enrollment = mean(enrollment)) 

ggplot(sch_sum, aes(x = year, y = sch, color = department, group = interaction(department, iteration))) + 
  geom_line(alpha=0.1) + 
  ggtitle("Simulated SCH Generated in Each Department") +
  guides(color = guide_legend(override.aes=list(alpha=1)))
```

Overall, we can look at the distributions of SCH generated by each department, which are shown in @fig-sch-gen-dept-dist. 

```{r}
#| label: fig-sch-gen-dept-dist
#| fig-width: 9
#| fig-height: 3
#| out-width: 100%
#| fig-cap: Distribution of SCH by department for each year. Years are shown as separate density curves, but the underlying functional mechanism is the same. 


ggplot(sch_sum, aes(x = sch, color = department, group = interaction(year, department))) + 
  geom_density() + 
  ggtitle("Simulated SCH Generated in Each Department") + facet_wrap(~department, scales = "free_x")
```


```{r}
#| label: fig-overall-sch-difference
#| fig-cap: Difference in total SCH generated across the 3 departments, 2024-2020. Because 2024 and 2020 are essentially drawn from the same distribution, our distribution is roughly symmetric around zero. 
#| fig-width: 6
#| fig-height: 4
#| out-width: "50%"
overall_5y_sum <- sch_results |>
  group_by(iteration, year) |>
  summarize(sch = sum(sch), enrollment = mean(enrollment)) |>
  summarize(total_sch_2024_2020 = sch[5]-sch[1])


ggplot(overall_5y_sum, aes(x = total_sch_2024_2020)) + 
  geom_density() + 
  ggtitle("Simulated Difference in Total University SCH Generated, 2024-2020")
```

```{r}
#| label: fig-dept-sch-difference
#| fig-cap: Difference in SCH generated for each of 3 departments as well as for the university overall, 2024-2020. Because 2024 and 2020 are essentially drawn from the same distribution, each distribution is roughly symmetric around zero, with  increasing variance as the number of faculty (and number of SCH generated) increase. 
#| fig-width: 8
#| fig-height: 4
#| out-width: "50%"

sch_5y_sum <- sch_sum |>
  left_join(overall_5y_sum) |>
  group_by(iteration, department) |>
  arrange(year) |>
  summarize(total_sch_2024_2020 = unique(total_sch_2024_2020), 
            sch_2024_2020 = sch[5]-sch[1], 
            share_growth_2024_2020 = sch_2024_2020/total_sch_2024_2020) |>
  ungroup() |>
  bind_rows(mutate(overall_5y_sum, sch_2024_2020 = total_sch_2024_2020, department = "Overall")) |>
  arrange(iteration) |>
  mutate(department = factor(department, levels = c("Small", "Med", "Large", "Overall"), ordered = T))

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

ggplot(sch_5y_sum, aes(x = sch_2024_2020, color = department)) + 
  geom_density() + 
  scale_color_manual("Unit", values = c("Small" = "#F8766D", "Med" = "#00ba38", "Large" = "#619CFF", "Overall" = "black")) + 
  ggtitle("Simulated Difference in Department SCH Generated, 2024-2020")
```


What do the ratios of department SCH to institutional SCH look like? 

